Elo World, a framework for \\ benchmarking weak chess \\ engines

### 2019-03-14 trunk/chess/papers/weak.tex
Writing on the plane, describing some more players

\describeplayer{random} We must begin with the most canonical of all
strategies: Choosing a legal move uniformly at random. This is a
lovely choice for Elo World, for several reasons: It is simple to
describe. It is clearly canonical, in that anyone undertaking a
similar project would come up with the same thing. It is capable of
producing any sequence of moves, and thus completely spans the gamut
from the worst possible player to the best. If we run the tournament
long enough, it will eventually draw games even against a hypothetical
perfect opponent, a sort of Boltzmann Brilliancy. Note that this
strategy actually does keep state (the pseudorandom pool), despite the
admonition above. We can see this as not really state but a simulation
of an external source of ``true'' randomness. Most other players
fall back on making random moves to break ties or when their primary
strategy does not apply. \canonical

@@

\describeplayer{first\_move} Make the lexicographically first legal
move. The moves are ordered as $\langle\!\!$ \verb+src_row+,
\verb+src_col+, \verb+dst_row+, \verb+dst_col+,
\verb+promote_to+ $\!\!\rangle$ for white (rank $1$ is row $0$) and the rows
are reversed for black to make the strategy symmetric. Tends to
produce draws (by repetition), because knights and rooks can often
move back and forth on the first few files. \deterministic

@@

@@

@@

@@

@@

\describeplayer{worstfish}
\describeplayer{safe}
\describeplayer{popular}
\describeplayer{dangerous}
\describeplayer{rare}
\describeplayer{survivalist}
\describeplayer{sym\_mirror\_x}
\describeplayer{sym\_180}
\describeplayer{sym\_mirror\_y}
\describeplayer{suicide\_king}
\describeplayer{fatalist}
\describeplayer{blind\_yolo}
\describeplayer{cccp}
\describeplayer{equalizer}
\describeplayer{blind\_kings}
\describeplayer{min\_oppt\_moves}
\describeplayer{chessmaster.nes\_lv1}
\describeplayer{stockfish1m\_r32768}
\describeplayer{chessmaster.nes\_lv2}
\describeplayer{stockfish1m\_r16384}
\describeplayer{stockfish0}
\describeplayer{stockfish5}
\describeplayer{stockfish1m\_r8192}
\describeplayer{topple10k}
\describeplayer{stockfish15}
\describeplayer{stockfish10}
\describeplayer{stockfish1m\_r4096}
\describeplayer{stockfish20}
\describeplayer{topple1m}
\describeplayer{stockfish1m\_r2048}
\describeplayer{stockfish1m\_r1024}
\describeplayer{stockfish1m\_r512}
\describeplayer{stockfish1m\_r256}
\describeplayer{stockfish1m\_r64}
\describeplayer{stockfish1m}
\describeplayer{stockfish1m\_r128}

@@

% worstfish (black) somehow beating suicide king
% white forces the mate. Note that worstfish
% does assume its opponent will pick good moves...
1. Nf3 g5 2. b4 Na6 3. Nxg5 Nc5 4. b5 f6
5. c4 Na6 6. Nf7 Rb8 7. d3 h6 8. Kd2 Nb4
9. Kc3 Rh7 10. Kd4 Rg7 11. Kc5 Nd5 12. Nxh6 Nb4
13. Na3 Rg3 14. Qc2 Bg7 15. f4 a6 16. Qb1 Bh8
17. Qb2 f5 18. Qf6 Rxg2 19. Qh4 Rxe2 20. Bg2 e5
21. Bd2 Bf6 22. Nf7 Bh8 23. Qf2 c6 24. Kd6 Re4
25. h4 Qc7+ 26. Kxc7 Nxa2 27. Raf1 c5 28. Nh6 Ke7
29. Qd4 Ke6 30. Rc1 Bg7 31. Be1 Ra8 32. Rc2 Nc3
33. Ra2 Ne7 34. Qxe4 Bf8 35. Kd8 Nxb5 36. Qxf5+ Kd6
37. Qxd7+ Bxd7++

>>>>

\describeplayer{random} We must begin with the most canonical of all
strategies: Choosing a legal move uniformly at random. This is a
lovely choice for Elo World, for several reasons: It is simple to
describe. It is clearly canonical, in that anyone undertaking a
similar project would come up with the same thing. It is capable of
producing any sequence of moves, and thus completely spans the gamut
from the worst possible player to the best. If we run the tournament
long enough, it will eventually at least draw games even against a
hypothetical perfect opponent, a sort of Boltzmann Brilliancy. Note
that this strategy actually does keep state (the pseudorandom pool),
despite the admonition above. We can see this as not really state but
a simulation of an external source of ``true'' randomness. Most other
players fall back on making random moves to break ties or when their
primary strategy does not apply. \canonical

@@

\describeplayer{first\_move} Make the lexicographically first legal
move. The moves are ordered as $\langle\!$~\verb+src_row+,
\verb+src_col+, \verb+dst_row+, \verb+dst_col+,
\verb+promote_to+~$\!\rangle$ for white (rank $1$ is row $0$) and the rows
are reversed for black to make the strategy symmetric. Tends to
produce draws (by repetition), because knights and rooks can often
move back and forth on the first few files. \deterministic

@@

% TODO: Cyrillic tag or toast at the end
\describeplayer{cccp} Prioritize moves that Checkmate, Check, Capture
or Push, in that order. Push means to move pieces as deep as possible
into enemy territory (without any regard for their safety). Ties are
broken deterministically by the source/destination squares, so this
one is technically asymmetric. \deterministic \asymmetric

\describeplayer{suicide\_king} Take a random move that minimizes the
distance between the two kings. Putting one's king out in the open is
very unprincipled (Figure~\ref{fig:cccpsuicide}), but it does produce
enough pressure to win against unambitious opponents.

\begin{figure}[ht]
  \chessboard[setfen=r1bq1bnr/pppp1p1p/2PQ4/8/2k6/8/PPP1PPPP/RN2KBNR b KQ - 0 7,showmover=false]
\caption{\player{suicide\_king} (black) dramatically failing against
  \player{cccp}'s slightly more principled play, after 1.~d4
  g5 2.~Bxg5 Nc6 3.~Bxe7 Kxe7 4.~d5 Kd6 5.~dxc6+ Kc5 6.~Qd6+ Kc4.
  White delivers a discovered mate with 7.~e4++.} \label{fig:cccpsuicide}
\end{figure}

% XXX include examples of symmetric players, as they are aesthetic.
\describeplayer{sym\_mirror\_y} As white, try to maximize symmetry
when flipping vertically. Zero penalty for opposing e.g.~a \King\, with
a \king, but a small penalty when the pieces are not the same type,
and a larger penalty when they are not opposite colors. The starting
position is already symmetric this way, so this usually has the effect
of copying the opponent's moves when possible. As ``copy your
opponent's moves'' is a common (but underspecified) strategy
discovered by many children, this player is close to being canonical.
However, it admits a bit too much arbitrary choice in the penalties
assigned.

\describeplayer{sym\_mirror\_x} As \player{sym\_mirror\_y}, but
maximize symmetry when flipping horizontally. This does not make much
chess sense, but can produce aesthetic arrangements
(Figure~\ref{fig:XXX}).

% XXX show self play perhaps?
\describeplayer{sym\_180} As \player{sym\_mirror\_y}, but maximize
symmetry under 180 degree
% XXX latex symbol for ``degrees''
rotation of the board. An emergent priority is
to ``castle'' with the king and queen to ``fix'' them.

\describeplayer{min\_oppt\_moves} Take a move that minimizes the
number of resulting legal moves for the opponent, breaking ties
randomly. This is a beautifully simple approach that generalizes many
chess principles: Occupying space reduces the destination squares
available to the opponent; capturing their pieces reduces the number
of their pieces that they can move; pinning pieces or checking the
king further reduces the legal moves; and mating the opponent is the
best possible move.\footnote{However note that it does not distinguish
  checkmate and stalemate, despite these having very different
  results.} Among the players in the paper, this one is Pareto
efficient in terms of its simplicity and effectiveness. \canonical

% XXX cite blog comment?
\describeplayer{equalizer} Prefer to move a piece that has been
moved the fewest times, and then prefer moving to a square that
has been visited the fewest times. Castling counts as moving
both the king and rook, and visiting both destination squares.
This is the first strategy described that keeps meaningful state.
\canonical \stateful

@@

If we keep state, then we can track the location of each piece as it
moves around the board (allowing us to distinguish the two rooks, or
follow a pawn as it promotes). We can then use statistics on the
average fates of each piece over hundreds of millions of games to
guide the moves. These statistics give us, for each piece (e.g.~the c2
pawn) and square, how likely it is to end the game on that square, and
how likely it is to be alive or dead there when the game
ends~\cite{survival}.

\describeplayer{safe} This strategy moves pieces towards squares where
they are likely to end the game alive. For this strategy and several
others, simply moving to maximize this score (e.g.~its sum or product
over all pieces) is very boring, since the score is almost always
maximized in the starting position. So this strategy actually makes
each move randomly, weighted by the total score of the resulting
positions. The scores are normalized (with the lowest-scoring move
receiving weight 0.0 and the highest 1.0) and then sampled according
to these weights. Without normalization, the play is almost identical
to uniformly random, since the weights of the resulting positions tend
to be very similar (dominated by the many pieces that don't move). But
it's pretty arbitrary.

\describeplayer{popular} Like \player{safe}, but the score for a
piece/square pair is the probability that the piece ends on that
square, whether it lives or dies. This player likes to follow the
crowd!

\describeplayer{dangerous} The dual of \player{safe}; the score is the
probability that the piece dies on that square. Note that a king is
said to die if it is checkmated or his side resigns. This player likes
to live life on the edge!

\describeplayer{rare} The dual of \player{dangerous}; the score
is one minus the probability of ending the game on that square. This
player has a thirst for adventure!

\describeplayer{survivalist} Like the above, but the score is the
ratio of the survival and death probabilities on the square. In the
data set, every piece ends alive or dead in every square (except
for the bishops, which can only legally occupy squares of their
color) at least 1000 times, so each ratio is defined. Here, the
sums of ratios have plenty of variability, and the highest
ratios are not so often on the starting squares. So with this
strategy, we simply do a weighted sample from the moves according
to their (non-normalized) scores.

\describeplayer{fatalist} The dual of \player{survivalist}; the score is
the ratio of the death and survival probability on the square. This
player knows that even if you win, you just have to keep playing over
and over again, so you might as well get it over with!

@@

Of course, people have been making more serious attempts at automating
chess since before computers, and there are thousands of chess engines
out there. We include a few in here to represent the high end of
skill, and to make sure that weaker players are evaluated somewhat in
terms of their ability to play chess proper, not just beat other weak
players.

% XXX citations for stockfish
\describeplayer{stockfish0} Stockfish is probably the strongest
open-source chess engine (or even publicly available engine); at full
strength its play is estimated to be around 3500 on the FIDE scale.
Aside from being quite machine dependent (it can search more moves in
a given amount of time when it has a fast CPU with many cores), there
are many options to fiddle with. Stockfish can use both opening books
and endgame tables; neither is used here. It also has a ``difficulty''
setting, which is set here to 0. Stockfish is run in a separate
process, and the board is reset before each move, but I am not
extremely hygienic about flushing e.g.~internal hash tables between
moves. One consequence of this attempt at statelessness is that
Stockfish sometimes walks into a draw by repetition in positions
where it would be able to win, because it doesn't know that it is
repeating positions.

% TODO: Does lichess publish the difficulty settings ? Nice to compare
% to my subjective skill level
\describeplayer{stockfish5} Stockfish as above, but set to difficulty
5.
\describeplayer{stockfish10} Same, at difficulty 10.
\describeplayer{stockfish15} Same, at difficulty 15.
\describeplayer{stockfish20} And difficulty 20, the maximum. Even
at this difficulty, Stockfish produces moves basically instantaneously.

\describeplayer{stockfish1m} As expected, the engine's performance
increases steadily as the difficulty setting increases (without
apparently affecting the time to make moves). I don't know what it's
doing with these settings. The true way to unleash chess engines is to
give them a lot of CPU and memory to search. Since the tournament is
run simultaneously across about 60 threads\footnote{ The computer is
  the completely egregious AMD 2990WX ``Threadripper~2'', which has 32
  cores (for 64 hardware threads) and 250 Watts of power dissipation
  at load. The torture of this CPU was part of the impetus for the
  paper.} using dozens of gigabytes of memory, and sometimes I would
play {\it Hollow Knight} while it ran, I wanted to avoid having the
performance be dependent on the scheduling environment. So here,
Stockfish is given a ``node'' budget of 1~million, hence {\sf 1m}. It
takes about one second per move when running alone, and is easily the
strongest player evaluated.

\begin{figure}[ht]
% worstfish (black) somehow beating suicide king
% white forces the mate. Note that worstfish
% does assume its opponent will pick good moves...
1. Nf3 g5 2. b4 Na6 3. Nxg5 Nc5 4. b5 f6
5. c4 Na6 6. Nf7 Rb8 7. d3 h6 8. Kd2 Nb4
9. Kc3 Rh7 10. Kd4 Rg7 11. Kc5 Nd5 12. Nxh6 Nb4
13. Na3 Rg3 14. Qc2 Bg7 15. f4 a6 16. Qb1 Bh8
17. Qb2 f5 18. Qf6 Rxg2 19. Qh4 Rxe2 20. Bg2 e5
21. Bd2 Bf6 22. Nf7 Bh8 23. Qf2 c6 24. Kd6 Re4
25. h4 Qc7+ 26. Kxc7 Nxa2 27. Raf1 c5 28. Nh6 Ke7
29. Qd4 Ke6 30. Rc1 Bg7 31. Be1 Ra8 32. Rc2 Nc3
33. Ra2 Ne7 34. Qxe4 Bf8 35. Kd8 Nxb5 36. Qxf5+ Kd6
37. Qxd7+ Bxd7++
\caption{TODO} \label{fig:worstfish}
\end{figure}

\describeplayer{worstfish} On the other hand, a strong chess engine
can also be used to play badly. When playing as white, for each legal
move, I ask Stockfish (configured as \player{stockfish0}) to evaluate
the resulting position from black's perspective.\footnote{By asking it
  to make a move, which also returns the evaluation of its move. The
  UCI protocol does not seem to offer a way to evaluate a position
  directly.} I then choose the move that produces the best position
for black. This is easily the worst player evaluated, but it is not
hard to imagine ways it could be worse. Indeed, playing for a loss is
a common twist on chess ({\it Losing
  Chess}~\cite{losingchesswikipedia}), although the rules of this
variant require that you capture a piece if you are able to (without
this constraint it is fairly boring). So, engines that support this
variant cannot be directly applied. We could use stronger settings of
Stockfish, but since it already invokes Stockfish for each legal move,
it is also one of the slowest players. Like Stockfish, it occasionally
blunders an otherwise losing position into a draw by repetition. But
most importantly, its search strategy when evaluating positions is not
applying the correct logic ($\alpha--\beta$ pruning); it assumes that
its opponent will choose strong moves, and that it will itself play
strong moves in future plies. As a result, it sometimes allows the
opponent to create a situation where \player{worstfish} is forced to
checkmate its opponent (Figure~\ref{fig:worstfish}).

\describeplayer{topple10k} Topple~\cite{topple} is another strong,
open source engine, provided to keep Stockfish on its toes. Here,
its node budget is 10,000. \traditional

\describeplayer{topple1m} Topple with a node budget of 1 million,
as \player{stockfish1m}. Stockfish almost always wins, though it
is not clear whether the budgets are actually comparable. \traditional

\describeplayer{chessmaster.nes\_lv1}
\describeplayer{chessmaster.nes\_lv2} \deterministic

@@

\describeplayer{blind\_yolo}
\describeplayer{blind\_kings}
% \describeplayer{blind\_spycheck} (described in paper, not yet run)

@@

\describeplayer{stockfish1m\_r32768}
\describeplayer{stockfish1m\_r16384}
\describeplayer{stockfish1m\_r8192}
\describeplayer{stockfish1m\_r4096}
\describeplayer{stockfish1m\_r2048}
\describeplayer{stockfish1m\_r1024}
\describeplayer{stockfish1m\_r512}
\describeplayer{stockfish1m\_r256}
\describeplayer{stockfish1m\_r64}
\describeplayer{stockfish1m\_r128}

@@
